Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-04 paper/revision2/ResponseToReviewers.tex

\documentclass[12pt]{article}
\usepackage[a4paper,top=2cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{float}
\usepackage{subfigure}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{color}
\usepackage{gb4e}
\noautomath

%\usepackage[natbib=true,maxcitenames=2,bibstyle=authoryear, style=apa]{biblatex}
\usepackage{natbib}
\bibliography{interferencebib}
\usepackage{lineno,xcolor,clipboard,graphicx}

\usepackage{csquotes}
%\DeclareLanguageMapping{american}{american-apa}

@@

Thank you for submitting your revised manuscript to the Journal of Memory and Language. This has now been reviewed by the same three experts who reviewed the previous submission, and you can find their comments below.

You will see from these that all three reviewers appreciate the thoroughness with which you responded to the issues raised on the initial submission. Reviewers 1 and 3 raise some remaining points but recommend acceptance, while Reviewer 2 is less persuaded that your revisions have adequately addressed their concerns. Given this set of reviews, I have come to a `revise' decision because I would like you to have one more go at addressing the remaining points made by all three reviewers.

I do not anticipate sending out the next version for review again, but rather anticipate being in a position to accept it at that point. However, it's important to note that that is not guaranteed, and one reason I have come to a `revise' rather than `accept subject to minor revisions' is that, like Reviewer 1, I was unable to access your data via your OSF link. \textbf{It is very important to us at JML that data are readily accessible, and that accessibility needs to be in place before any potential acceptance decision}. 
''
\end{quote}

\subsubsection*{Response to E.1}
We have made the OSF repository public and updated the link in the manuscript. This should grant general accessibility.

\textcolor{red}{wait for ok from shravan}

@@

\textbf{Reviewer 2 and 3's clear view that you still overstate the implications of your findings in places}.''
\end{quote}

\subsubsection*{Response to E.2}
\textcolor{red}{add *in this design* whenever we say that there is no syn interference}

\subsection*{Comment E.3}
\begin{quote}
``\textbf{Reviewer 2's additional point 1 that asks whether you have changed that particular analysis}.''
\end{quote}

\subsubsection*{Response to E.3}
Yes, we changed all analyses in the first revision. As we had explained in the response to reviewers on August 29, 2024 (specifically in the response letter and our responses to the comments E.2 and R3.5), the priors were changed from directional priors in the original version to symmetrical priors in the revised version.

@@

I do have some suggestions. In my view these are only minor. They mainly concern just the way things are presented or discussed. I list my comments by page.

p. 16: osf data etc.

I wanted to check those, but \textbf{when I clicked the link, it said I don't have permissions and that I need to request access}.''
\end{quote}

\subsubsection*{Response to R1.1}
We have made the OSF repository public and updated the link in the manuscript. This should grant general accessibility.

@@

\subsection*{Comment R1.2}

\begin{quote}
``p. 17, SPR experiment

\textbf{Did the analysis include any cut-off point for reading times to exclude data? I could not find this in the manuscript. If it did, could this information be included?} If not, I would strongly suggest to use this. In my own experience with Prolific, I noticed the following very common pattern: people basically do not read around 10-20\% of experimental items - that is, for quite a significant subset of participants I noticed that they have super fast reading times on some subset of items, most likely because they just rushed through those items by just pressing the space bar and keeping it pressed. This is sometimes followed by a long waiting time at some other point, so those participants would not be outliers when we check the total time in the experiment. These people and their data would not be excluded by checking their answers to comprehension questions, especially if only a small subset of items has comprehension questions, like one third, as in this experiment (basically, they could still easily pass with around 90\% success rate). So if this was not checked, it would be good to check and \textbf{consider removing RTs based on small/large cut-off points and to see whether that affects the analysis. If this was done already, it would be good to report more details, in particular, the cut-off points and also the amount of data that was removed this way}.''
\end{quote}

\subsubsection*{Response to R1.2}
We excluded reading times below 150 ms and above 3000 ms. This affected 4.9\,\% of the data. We have added this information to the manuscript on page \pageref{trim}:

\begin{quote}
    \Paste{trim}
\end{quote}

@@

\subsection*{Comment R1.3}
\begin{quote}
``p. 36: ``Regarding the possibility that sentence structure confounds caused the effects in the pre-critical region(.)"

\textbf{I wonder whether it would make sense to briefly summarize what the sentence structure confound should be.} Otherwise the whole paragraph is hard to follow for readers who did not read or do not remember Mertzen et al. (2023).''
\end{quote}
\subsubsection*{Response to R1.3}
We have added the following text to the manuscript on page \pageref{confound} (above the sentence which was quoted by the reviewer):

\begin{quote}
    \Paste{confound}
\end{quote}

@@

\subsubsection*{Response to R1.5}
\textcolor{red}{Opened issue and assigned to Himanshu}

\subsection*{Comment R1.6}

\begin{quote}
``p. 61, \textbf{Figure 14: one value is outside the scale of the bottom left graph. If possible, it would be good to make it visible}.''
\end{quote}

\subsubsection*{Response to R1.6}
We have changed the y-axis limits to include the mean estimate of the respective data point (VD07.E2, regression path duration). For convenience, we paste the revised Figure here.

\setcounter{figure}{13}
\begin{sidewaysfigure}[htpb]
    \caption{A) Reading time differences in the regions of interest from the present and previous studies using the design by \cite{vandyke07}. VD07.E1, VD07.E2 and VD07.E3 stand for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiments 1-3. The intervals for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) estimates are 95\% confidence intervals. M23.E and M23.G stand for \citeauthor{mertzen}'s (\citeyear{mertzen}) English and German experiment, respectively. The intervals for their study and the present study are Bayesian 95\% credible intervals. There are less estimates for the pre-critical region because \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 1 and 2 did not have a pre-critical region. B) Bayes factors for the effects of interest in the present and previous studies under prior Normal(0,0.5).}
    \label{fig:previous_vs_present}
    \centering
    \includegraphics[width=\linewidth]{present_previous_allplots.jpg}
\end{sidewaysfigure}
\clearpage

@@

\subsection*{Comment R2.6}
\begin{quote}
``Additional points:

1. Original version: ``By contrast, the syntactic interference effect had only anecdotal evidence under the two narrow priors (Normal-(0, 0.1): BF10 = 2.5, Normal-(0, 0.5): BF10 = 1.2). There was evidence against syntactic interference under wider priors (Normal-(0, 1): BF10 = 0.66, Normal-(0, 5): BF10 = 0.14). What this means is that, in our data, only if we assume a priori that the effect size is relatively small, there is very weak evidence in favor of syntactic interference."

Revised version: ``By contrast, Bayes factors provided either no evidence for or even evidence against syntactic interference in both spatiotemporal windows (BF10 $<$ 1.2). Similarly, Bayes factors provided either no evidence for or even evidence against the interaction in both spatio-temporal windows (BF10 $<$ 1.2)."

\textbf{Did the authors change the analyses? Why what was reported as anecdotal evidence under narrow priors is now reported as no evidence or evidence against?}''
\end{quote}
\subsubsection*{Response to R2.6}
Yes, the priors were changed in the 1st revision (directional priors in the original version and symmetrical priors in the revised version, see also our Response to E.3). This resulted in small numerical changes of the results, e.g., BF10 = 2.5 in the original analysis decreased to BF10 = 1.2 in the revised analysis. Furthermore in the original version of the manuscript, we were inconsistent in our description of small BFs, i.e., BF10 = 1.2. Since the first revision and in line with \cite{jeffreys1998theory} and \cite{lee2014bayesian}, we now consistently interpret such small BFs as providing no evidence.

@@

\subsection*{Comment R3.3}
\begin{quote}
``Finally, and in a similar vein, I raised questions about whether there might be other reasons for the selective interference effects found in this study. Maybe a different mechanism could capture the special status of the subject-of-the-same-clause, or maybe some specific property of the current experimental materials could be responsible for the lack of interference from other subjects. The authors seem skeptical of the first of these, and they seem more sympathetic to the second. \textbf{This is all reasonable enough, and their text conveys some caution. The abstract, on the other hand, is rather more confident.} I am sympathetic to the authors' preferred conclusion. But I am less confident than they are about how well justified it is based on the current findings. This has nothing to do with the quantitative wizardry that the authors display in the paper. It's all about the issue of generalizing from a single sentence configuration (that participants saw in 60\% of trials in the study).''
\end{quote}

\subsubsection*{Response to R3.3}
We have revised abstract to emphasize that our findings only apply to the present design. We paste the revised abstract here:

\begin{quote}
    \Paste{abstract}
\end{quote}

### 2024-11-04 paper/revision2/elsarticle-template-harv.tex
Add files via upload

\usepackage{bigdelim}
\usepackage{hyperref}

@@

\begin{abstract}
%% Text of abstract
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that, in our data, although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.
\end{abstract}

@@

Reading times across the whole sentence are shown in Figure \ref{fig:whole_sentence} A and B, separately for high and low syntactic interference because these conditions had partially different sentence structures.

@@

\cite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. They discussed several alternative explanations for their observed effects. In addition to encoding interference, \cite{mertzen} proposed three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions, and predictive processing effects. Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading.

@@

For their recent simulations, \cite{mertzen} created a version of the Lewis and Vasishth (2005) model that assumes that in the \cite{vandyke07} design, three cues are used at the verb: \{$\pm$ animacy\}, \{$\pm$ grammatical subject\} and \{$\pm$ same clause\}. The \{$\pm$ same clause\} cue matches only with nouns that occurred in the same clause as the verb. ``The addition of the \{$\pm$ same clause\} cue [was] necessary to identify the correct subject'' and as a means for the model ``to achieve correct retrieval'' in the condition with high syntactic and high semantic interference \citep[][p. 14]{mertzen}. This model predicted approximately equally sized effects of syntactic and semantic interference. Although it is clear that our ERP data does not support equally sized effects of syntactic and semantic interference, we decided to include their version of the model here because it has previously been claimed to be a good model for the present design. We have implemented two alternative versions of the model. Given that in our ERP data, there was no evidence for syntactic interference or depending on the a priori assumed effect size even evidence against it, it is plausible that, at least in this experimental design and language, syntactic interference plays little to no role. We speculate that this could be due to the parser tracking hierarchical structure and that this hierarchical information is used when searching for a noun in memory -- even more so than suggested by \cite{mertzen}. \label{clause_tracker}{So, we implemented a model which puts even more emphasis on hierarchical syntactic structure and thereby basically eliminates syntactic interference. Specifically, the relevant syntactic cues that might be used at the verb are not \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} -- as implemented by \cite{mertzen} -- but rather a composite cue \{$\pm$ subject-in-same-clause\}. As already noted by \cite{mertzen}, structural information like [+ subject of clause$_i$] is not an inherent feature like for example [+ animate], but within in the cue-based retrieval framework, as a simplification, we assume that such a feature is encoded into memory and can consequently be used to match the retrieval cue \{$\pm$ subject of clause$_i$\}. This idea goes back to an assumption in the Lewis and Vasishth (2005) model by making the model ``sensitive to whether or not it is parsing an embedded clause'' \citep[][p.388]{Lewis2005} and implies a simple clause tracking mechanism. The resulting model assumes two retrieval cues: \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. This proposal is furthermore in the spirit of the \cite{dillon2013} and \cite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval. In the present case, it is possible that the syntactically relevant noun is only the one in the same clause as the verb. {\Copy{cite_kush}{For more discussion of the role of structural retrieval cues, see \textcolor{blue}{\citet{kush2015relation}}, \citet{franck2020hierarchical} and \citet{arnett2017subject}.}}\label{cite_kush} The proposed model here using two cues predicts no syntactic interference and no interaction, but does predict semantic interference effects. Lastly, we implemented a model with one cue which unambiguously matches the subject. This cue could be the composite cue \{$\pm$ animate-subject-in-same-clause\} or it could be even more specific. This model naturally does not predict any interference effects. It can be regarded the null model or no-interference model.

>>>>

\usepackage{bigdelim}
\usepackage{hyperref}
\usepackage{soul}

@@

\begin{abstract}
%% Text of abstract
\Copy{abstract}{
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. \textcolor{blue}{In this design, the syntactic cue \{+subject\} and the semantic cue \{+animate\} are manipulated.}
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, \textcolor{blue}{semantic interference due to the semantic cue \{+animate\}} caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to \textcolor{blue}{the syntactic cue \{+subject\}}. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference \textcolor{blue}{in the present design} if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that, in our data, although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects \textcolor{blue}{in sentence configurations like the ones investigated in the present study}.
}
\end{abstract}

@@

\Copy{trim}{{\textcolor{blue}{Reading times below 150 ms and above 3000 ms were excluded before analysis (and visualization) because these overly short or long reaction times were likely caused by inattentive reading. The procedure excluded 4.9\,\% of the data.}}}\label{trim} Reading times across the whole sentence are shown in Figure \ref{fig:whole_sentence} A and B, separately for high and low syntactic interference because these conditions had partially different sentence structures.

@@

\cite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. They discussed several alternative explanations for their observed effects. {\Copy{confound}{In addition to encoding interference, \cite{mertzen} proposed three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions \textcolor{blue}{(two vs.\ one embedded clause between subject and verb in the high vs.\ low syntactic interference conditions)}, and predictive processing effects.}}\label{confound} Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading.

@@

For their recent simulations, \cite{mertzen} created a version of the Lewis and Vasishth (2005) model that assumes that in the \cite{vandyke07} design, three cues are used at the verb: \{$\pm$ animacy\}, \{$\pm$ grammatical subject\} and \{$\pm$ same clause\}. The \{$\pm$ same clause\} cue matches only with nouns that occurred in the same clause as the verb. ``The addition of the \{$\pm$ same clause\} cue [was] necessary to identify the correct subject'' and as a means for the model ``to achieve correct retrieval'' in the condition with high syntactic and high semantic interference \citep[][p. 14]{mertzen}. This model predicted approximately equally sized effects of syntactic and semantic interference. Although it is clear that our ERP data does not support equally sized effects of syntactic and semantic interference, we decided to include their version of the model here because it has previously been claimed to be a good model for the present design. We have implemented two alternative versions of the model. Given that in our ERP data, there was no evidence for syntactic interference or depending on the a priori assumed effect size even evidence against it, it is plausible that, at least in this experimental design and language, syntactic interference plays little to no role. We speculate that this could be due to the parser tracking hierarchical structure and that this hierarchical information is used when searching for a noun in memory -- even more so than suggested by \cite{mertzen}. \label{clause_tracker}{So, we implemented a model which puts even more emphasis on hierarchical syntactic structure and thereby basically eliminates syntactic interference. Specifically, the relevant syntactic cues that might be used at the verb are not \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} -- as implemented by \cite{mertzen} -- but rather a composite cue \{$\pm$ subject-in-same-clause\}. As already noted by \cite{mertzen}, structural information like [+ subject of clause$_i$] is not an inherent feature like for example [+ animate], but within in the cue-based retrieval framework, as a simplification, we assume that such a feature is encoded into memory and can consequently be used to match the retrieval cue \{$\pm$ subject of clause$_i$\}. This idea goes back to an assumption in the Lewis and Vasishth (2005) model by making the model ``sensitive to whether or not it is parsing an embedded clause'' \citep[][p.388]{Lewis2005} and implies a simple clause tracking mechanism. The resulting model assumes two retrieval cues: \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. This proposal is furthermore in the spirit of the \cite{dillon2013} and \cite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval. In the present case, it is possible that the syntactically relevant noun is only the one in the same clause as the verb. {\Copy{cite_kush}{For more discussion of the role of structural retrieval cues, see \textcolor{blue}{\cite{kush2015relation}}, \cite{franck2020hierarchical} and \cite{arnett2017subject}.}}\label{cite_kush} The proposed model here using two cues predicts no syntactic interference and no interaction, but does predict semantic interference effects. Lastly, we implemented a model with one cue which unambiguously matches the subject. This cue could be the composite cue \{$\pm$ animate-subject-in-same-clause\} or it could be even more specific. This model naturally does not predict any interference effects. It can be regarded the null model or no-interference model.
